### 2025-07-15 manuscript/main.tex
type axoim for expressions no array yet
types and semantic done
custom services intro, scope computation
control flow - data flow; update architecture

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage[outputdir=out, newfloat]{minted}
\usepackage{multicol}
\usepackage{fontspec}

@@

\usepackage{forest}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage[colorlinks,linkcolor=blue,urlcolor=cyan]{hyperref}
\usepackage{caption}
\usepackage{titlesec}
\titleformat{\chapter}[hang]{\normalfont\huge\bfseries}{}{0pt}{}
\titlespacing*{\chapter}{0pt}{1.5em}{1em}
%=== Metadata ===
\newcommand{\thesistitle}{Your Thesis Title Here}
\newcommand{\authorname}{Rafael Tanzer}
\newcommand{\supervisor}{Michele Chiaria, Simone , Francesco Pontigia and Ezzio Bartocci}
\newcommand{\institution}{TU Vienna}
\newcommand{\department}{Cyber-Physical-Systems, ..., ...}
\newcommand{\logoimage}{graphics/logo.png} % Path to your logo image
\newcommand{\submissionmonth}{May}
\newcommand{\submissionyear}{2025}
%================

@@

@@

\chapter{Expanded Code Listings}
% Additional code excerpts.
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb}
\label{list:langium-grammar}

>>>>

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage[outputdir=out, newfloat]{minted}
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{
  arrows.meta,      % nicer arrow tips
  shapes.geometric, % rectangle, diamond, etc.
  positioning,      % easy relative placement
  chains,           % flow‐style “chain” syntax
  calc             % coordinate arithmetic
}
\tikzset{
  process/.style = {
    rectangle,
    draw,
    minimum width=3cm,
    minimum height=1cm,
    text centered,
    on chain
  },
  arrow/.style   = {
    -Stealth,
    thick
  },
  start chain    = going below,  % default chain direction
  node distance  = 8mm           % space between chain nodes
}
\usepackage{amsfonts}
\usepackage{fontspec}

@@

\usepackage{forest}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage[colorlinks,linkcolor=blue,urlcolor=cyan]{hyperref}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{titlesec}
\titleformat{\chapter}[hang]{\normalfont\huge\bfseries}{}{0pt}{}
\titlespacing*{\chapter}{0pt}{1.5em}{1em}
%=== Metadata ===
\newcommand{\thesistitle}{Your Thesis Title Here}
\newcommand{\authorname}{Rafael Tanzer}
\newcommand{\supervisor}{Michele Chiaria, Simone , Francesco Pontigia and Ezzio Bartocci}
\newcommand{\institution}{TU Vienna}
\newcommand{\department}{Cyber-Physical-Systems, ..., ...}
\newcommand{\logoimage}{graphics/logo.png} % Path to your logo image
\newcommand{\submissionmonth}{May}
\newcommand{\submissionyear}{2025}
%================

@@

\definecolor{lightgray}{gray}{0.97}

@@

\chapter{Expanded Code Listings}
% Additional code excerpts.
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb (LL(2) parsable)}
\label{list:langium-grammar}

### 2025-07-15 manuscript/part1.tex
control flow - data flow; update architecture

Created by the <Faculty ()CPS> - .... formal verification, model checking, temporal logics and automata - the MiniProb DSL provides a concise notation
for probabilistic process models\cite{guideMiniProb}. At present, MiniProb tooling is limited to a Haskell-based parser, yielding a tedious workflow: source code is written
in a generic editor and syntax errors discovered one at a time as the parser aborts on its first failure. A modern editor extension - for example, a
Visual Studio Code plugin providing syntax highlighting and live diagnostics - would dramatically streamline development. Such an interface lowers the barrier to entry for
new users and also offers a familiar environment in which MiniProb programs can be authored and the further automatically converted into automata for formal verification \cite{2025modelcheckingprobabilisticoperator,POPACheck}.
In the context of cyber-physical systems - where interconnected devices underpin critical infrastructure - and the ever-growing complexity and threat landscape of cyberspace,
integrated verification support is essential to ensure system reliability, preempt security vulnerabilities, and uphold correctness guarantees under rigorous temporal and
probabilistic constraints.

>>>>

Created by the supervising team - whose research spans formal verification, model checking, temporal logics and automata - the MiniProb DSL provides a concise notation
for probabilistic process models\cite{guideMiniProb}. At present, MiniProb tooling is limited to a Haskell-based parser, yielding a tedious workflow: source code is written
in a generic editor and syntax errors discovered one at a time as the parser aborts on its first failure. A modern editor extension - for example, a
Visual Studio Code plugin providing syntax highlighting and live diagnostics - would dramatically streamline development. Such an interface lowers the barrier to entry for
new users and also offers a familiar environment in which MiniProb programs can be authored and the further automatically converted into automata for formal verification \cite{2025modelcheckingprobabilisticoperator,POPACheck}.
In the context of cyber-physical systems - where interconnected devices underpin critical infrastructure - and the ever-growing complexity and threat landscape of cyberspace,
integrated verification support is essential to ensure system reliability, preempt security vulnerabilities, and uphold correctness guarantees under rigorous temporal and
probabilistic constraints.

### 2025-07-15 manuscript/part2.tex
control flow - data flow; update architecture

\section{High-Level Architecture Diagram}
update diagram such that cusomt services also extends the lsp providers
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{graphics/langiumArchitecture.png}
  \caption{Architecture overview of a Langium project}
  \label{fig:langium-architecture}
\end{figure}

\section{Module Decomposition}

@@

\begin{description}
  \item[\code{language/}] Assembles the core of any Langium project. It includes the \code{.langium} file, in which the custom language grammar is defined, as well as
    the code that starts the Language Server with all necessary services bundled.
  \item[\code{cli/}] Provides functionality for CLI interactions, command parsing, and generators. The generators in this folder invoke processes that generate artifacts
    from the language and input files; they are typically used only offline and are called once rather than continuously via the CLI and support efficient pipelining.
  \item[\code{extension/}] Contains the VS Code extension bootstrap code needed to connect to the Language Server and utilize LSP support features.
    All mentioned folders except the one containing CLI logic are bundled into the extension itself.
  \item[\code{generated/}] Contains artifacts produced from the grammar definition in the \code{.langium} file. This includes TypeScript AST type definitions,
    which are essential for further analysis and validation, as well as the preliminary parsed grammar rules formatted for the Chevrotain engine to generate parsers.
  \item[\textsf{Services}] The Services in Langium are a collection of utility classes and helper functions offered to every component of a language project via
    dependency injection. They cover an extensive range of functionality, from LSP-specific features - such as providers for context-aware code completion, rich hover
    tooltips, signature help, go-to-definition and find-references - to core language utilities that expose parsed documents as both AST and CST models, manage workspace
    state and context information (including indexing, scoping, and name resolution), synchronize the in-memory document model and perform validation checks.
    Because the parts are wired up as an injectable service, precise consumptions of the capabilities needed is possible. Moreover, Langium enables developers to swap in
    custom implementations of services by extending the corresponding default classes, and these services have to be wired into the DI container manually \ref{sec}.
    By default, the validation logic is located in the \code{language/} directory; subsequently, I placed additional custom services there.
    However, as these services grow in both number and complexity, it may be more appropriate to organize them in a dedicated module or folder.
\end{description}

@@

\section{Data Flow and Control Flow}
control flow of lsp from file to lsp to editor

@@

The \code{FileImport} rule emulates the C preprocessor's "\#include" directive by requiring a literal match of the \texttt{\#include} keyword followed by a
file name token (terminal \code{STRING}). Generally during parsing, each line is evaluated either against these fixed string keywords or against any matching terminals (order of match).
As the file names have to be retrievable once it is time to compute the scope \ref{sec:customservices}, they are made accessible through the AST
, with each FileImport-node containing a \texttt{file} property that stores the corresponding value.

>>>>

\section{High-Level Architecture Diagram \& Module Decomposition}
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{graphics/langiumArchitecture.png}
  \caption{Architecture overview of a Langium project}
  \label{fig:langium-architecture}
\end{figure}

@@

\begin{description}
  \item[\code{language/}] Assembles the core of any Langium project. It includes the \code{.langium} file, in which the custom language grammar is defined, as well as
    the code that starts the Language Server with all necessary services bundled.
  \item[\code{cli/}] Provides functionality for CLI interactions, command parsing, and generators. The generators in this folder invoke processes that generate artifacts
    from the language and input files; they are typically used only offline and are called once rather than continuously via the CLI and support efficient pipelining.
  \item[\code{extension/}] Contains the VS Code extension bootstrap code needed to connect to the Language Server and utilize LSP support features.
    All mentioned folders except the one containing CLI logic are bundled into the extension itself.
  \item[\code{generated/}] Contains artifacts produced from the grammar definition in the \code{.langium} file. This includes TypeScript AST type definitions,
    which are essential for further analysis and validation, as well as the preliminary parsed grammar rules formatted for the Chevrotain engine to generate parsers.
  \item[\textsf{Services}] The Services in Langium are a collection of utility classes and helper functions offered to every component of a language project via
    dependency injection. They cover an extensive range of functionality, from LSP-specific features - such as providers for context-aware code completion, rich hover
    tooltips, signature help, go-to-definition and find-references - to core language utilities that expose parsed documents as both AST and CST models, manage workspace
    state and context information (including indexing, scoping, and name resolution), synchronize the in-memory document model and perform validation checks.
    Because the parts are wired up as an injectable service, precise consumptions of the capabilities needed is possible. Moreover, Langium enables developers to swap in
    custom implementations of services by extending the corresponding default classes, and these services have to be wired into the DI container manually \ref{sec:customservices}.
    By default, the validation logic is located in the \code{language/} directory; subsequently, I placed additional custom services there.
    However, as these services grow in both number and complexity, it may be more appropriate to organize them in a dedicated module or folder.
\end{description}

@@

\section{Data Flow and Control Flow}

\begin{tikzpicture}[node distance=8mm and 12mm]
  % Top row
  \node[process]      (A) {1. Document Change};
  \node[process,right=of A] (B) {2. Tokenization};
  \node[process,right=of B] (C) {3. Parsing};
  \node[process,right=of C] (D) {4. CST $\to$ AST};

  % Bottom row (starts under D, reverses direction)
  \node[process,below=of D] (E) {5. Linking \& Scoping};
  \node[process,left=of E]  (F) {6. Validation};
  \node[process,left=of F]  (G) {7. Diagnostics};
  \node[process,left=of G]  (H) {8. Semantic Highlighting};

   \foreach \src/\tgt in {A/B, B/C, C/D, D/E, E/F, F/G, G/H} {
    \draw[arrow] (\src) -- (\tgt);
  }
\end{tikzpicture}
\captionof{figure}{Flow chart of Langium's document process pipeline}
\vspace{0.5em}

The editor (via VS Code's LSP client) detects file opens or edits and sends an \code{onDidChangeTextDocument} notification to the server \cite{vscode-lsp}. Chevrotain's Lexer then performs lexical analysis,
breaking the raw character stream into a sequence of tokens \cite{chevrotain-docs}, which the Chevrotain parser consumes to build a Concrete Syntax Tree (CST). Langium's AST Builder walks the CST to produce
a higher-level Abstract Syntax Tree (AST) \cite{langium-doc-lifecycle}, after which the \code{ScopeProvider} and \code{Linker} resolve cross-references, the \code{ValidationRegistry} applies semantic rules
to generate diagnostics, and the server finally publishes diagnostics, semantic tokens, completions, hover information, and other language-service results back to the editor via the
Language Server Protocol.

\subsection*{Language Server Protocol: Server client interaction}

This illustration is a concise comprehension aid to show how the Language Server Protocol enables standardized communication between an editor (client) and a language server:
from the initialization handshake through document events, feature requests (completion, hover, definition, references, formatting, semantic tokens), and shutdown \cite{lspFeatures}.
By defining a common set of well-specified messages and payloads, LSP serves as a generally applicable protocol that any editor and any language server can adopt to instantly
share rich tooling capabilities without per-language or per-IDE custom integrations.

\begin{center}
  \begin{tikzpicture}[
  lifeline/.style={rectangle, draw, minimum width=3cm, text centered},
  message/.style={-{Stealth}, thick},
  msglabel/.style={font=\small, text width=5cm, align=center}
]
  % Lifeline heads
  \node[lifeline] (C) {Client};
  \node[lifeline,right=8cm of C] (S) {Server};

  % Lifelines
  \draw[dashed] (C.south) -- ++(0,-21*8mm-5mm);
  \draw[dashed] (S.south) -- ++(0,-21*8mm-5mm);

  \node at
    ($ (C.north)!.5!(S.north) + (0,5mm) $)
    {\texttt{(JSON-RPC)}};

 \foreach \i/\txt in {
    1/initialize,
    3/initialized (notification),
    4/textDocument\slash didOpen,
    5/textDocument\slash didChange,
    6/textDocument\slash completion,
    8/textDocument\slash hover,
    10/textDocument\slash definition,
    12/textDocument\slash references,
    14/textDocument\slash formatting,
    16/textDocument\slash semanticTokens,
    18/textDocument\slash didSave,
    19/textDocument\slash didClose,
    20/shutdown,
    21/exit
  }{
    \draw[message]
      ($ (C.south) + (0,-\i*8mm) $)
      -- 
      ($ (S.south) + (0,-\i*8mm) $)
      node[midway, above, msglabel] {\txt};
  }

  % Server→Client messages (labels below)
  \foreach \i/\txt in {
    2/initialize response,
    7/completion items,
    9/hover info,
    11/definition locations,
    13/reference locations,
    15/edits,
    17/token data
  }{
    \draw[message]
      ($ (S.south) + (0,-\i*8mm) $)
      -- 
      ($ (C.south) + (0,-\i*8mm) $)
      node[midway, above, msglabel] {\txt};
  }
\end{tikzpicture}
\end{center}

@@

The \code{FileImport} rule emulates the C preprocessor's "\#include" directive by requiring a literal match of the \code{#include} keyword followed by a
file name token (terminal \code{STRING}). Generally during parsing, each line is evaluated either against these fixed string keywords or against any matching terminals (order of match).
As the file names have to be retrievable once it is time to compute the scope \ref{sec:customservices}, they are made accessible through the AST
, with each FileImport-node containing a \code{file} property that stores the corresponding value.
